---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ebviz

<!-- badges: start -->
<!-- badges: end -->

The goal of `ebviz` is to assist authors of a Multiple Regression Poststratification
edited book in creating visualizations that are readable, cohesive, and accessible.

## Installation

You can install the `ebviz` package from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("j-effendy/ebviz")
```
## Example

These are basic plot examples of `theme_seagull()` and `theme_wombat()` using 
`mtcars` and `economics_long` dataset:

```{r theme, echo=FALSE, fig.width=10, fig.height=15}

library(ebviz)
library(ggplot2)
library(patchwork)

w1 <- ggplot(mtcars, aes(mpg,disp,color=factor(carb))) + 
  geom_point(size=3) + 
  labs(title="Scatter Plot",
       subtitle = "theme_wombat()",
       caption = "Data: mtcars") +
  theme_wombat()

w2 <- ggplot(mtcars, aes(factor(carb),fill=factor(carb))) + 
  geom_bar() + 
  scale_y_continuous(expand = c(0,0)) + 
  scale_x_discrete(expand = c(0,0)) +
  labs(title="Bar Plot",
       subtitle = "theme_wombat()",
       caption = "Data: mtcars") +
  theme_wombat()

w3 <- ggplot(economics_long, aes(date, value01, colour = variable)) +
  geom_line() +
   scale_y_continuous(expand = c(0,0)) + 
   scale_x_date(expand = c(0,0)) +
  labs(title = "Line Plot",
       subtitle = "theme_wombat()",
       caption = "Data: economics_long") +
  theme_wombat()

s1 <- ggplot(mtcars, aes(mpg,disp,color=factor(carb))) + 
  geom_point(size=3) + 
  labs(title="Scatter Plot",
       subtitle = "theme_seagull()",
       caption = "Data: mtcars") +
  theme_seagull()

s2 <- ggplot(mtcars, aes(factor(carb),fill=factor(carb))) + 
  geom_bar() + 
  scale_y_continuous(expand = c(0,0)) + 
  scale_x_discrete(expand = c(0,0)) +
  labs(title="Bar Plot",
       subtitle = "theme_seagull()",
       caption = "Data: mtcars") +
  theme_seagull()

s3 <- ggplot(economics_long, aes(date, value01, colour = variable)) +
  geom_line() +
  scale_y_continuous(expand = c(0,0)) + 
  scale_x_date(expand = c(0,0)) +
  labs(title = "Line Plot",
       subtitle = "theme_seagull()",
       caption = "Data: economics_long") +
  theme_seagull()

(s1 + w1) / (s2 + w2) / (s3 + w3)
```

These are basic plot examples of `ebviz` colour palettes using 
`mtcars` dataset:

```{r colour, echo=FALSE, fig.width=10, fig.height=15}
library(dataviz)
library(ggplot2)
library(patchwork)

p1 <- ggplot(mtcars, aes(mpg,disp,color=factor(vs))) +
        geom_point(size=3) +
        scale_colour_dataviz("quokka") +
        labs(title="Scatter Plot",
             subtitle = "Quokka Palette",
             caption = "Data: mtcars") +
        theme_wombat()

p2 <- ggplot(mtcars, aes(mpg,disp,color=factor(vs))) +
        geom_point(size=3) +
        scale_colour_dataviz("quoll") +
        labs(title="Scatter Plot",
               subtitle = "Quoll Palette",
               caption = "Data: mtcars") +
        theme_wombat()

p3 <- ggplot(mtcars, aes(mpg,disp,color=mpg)) +
        geom_point(size=3) +
        geom_point(size=3) +
        scale_colour_dataviz("seal") +
        labs(title="Scatter Plot",
                 subtitle = "Seal Palette",
                 caption = "Data: mtcars") +
        theme_wombat()

p4 <- ggplot(mtcars, aes(mpg,disp,color=mpg)) +
        geom_point(size=3) +
        geom_point(size=3) +
        scale_colour_dataviz("snake") +
        labs(title="Scatter Plot",
                 subtitle = "Snake Palette",
                 caption = "Data: mtcars") +
        theme_wombat()

p5 <- ggplot(mtcars, aes(mpg,disp,color=mpg)) +
        geom_point(size=3) +
        scale_colour_dataviz("dingo") +
        labs(title="Scatter Plot",
                 subtitle = "Dingo Palette",
                 caption = "Data: mtcars") +
        theme_wombat()

p6 <- ggplot(mtcars, aes(mpg,disp,color=mpg)) +
        geom_point(size=3) +
        scale_colour_dataviz("dolphin") +
        labs(title="Scatter Plot",
                 subtitle = "Dolphin Palette",
                 caption = "Data: mtcars") +
        theme_wombat()

(p1 + p2) / (p3 + p4) / (p5 + p6)
```
