---
title: "Project Report"
subtitle: "Automatic Visualisations Aesthetic Production for Multilevel Regression and Poststratification Book"
author: "Jimmy Effendy"
date: "13/10/2021"
output: bookdown::pdf_document2
bibliography: references.bib
csl: apa.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(ggplot2)
library(tidyverse)
library(dataviz)
library(colorspace)
library(patchwork)
library(pals)
```

\newpage

# Abstract

The work involved in the productions of visualizations in an edited book
is often laborious. With many authors involved, it is difficult to produce
an edited book that has a cohesive design. This project aims to reduce 
workflow required in producing visualizations that are readable, cohesive, and
accessible in a Multilevel Regression and Poststratification (MRP) edited book.
The project contributes to a ggplot theme and accessible colour schemes. The 
theme is developed based on Tufte's principle of data graphical excellence.
In addition, the project also develops two colour schemes for each type of
colour palettes: qualitative, sequential, and diverging palettes. These theme
and colour palettes are bundled together in an R package called `dataviz`.
An R package is used for this project as authors can easily install the package
and use its functions for their research.

# Introduction

This project relates to an edited book about Multilevel Regression and 
Poststratification (MRP). As is the case with other books, there will be
numerous visualizations that will be used within the book. These plots will need to be
readable, cohesive, and accessible. There is a group of authors (around one to 
five authors) who work together on a chapter. The editors then work to combine 
all the chapters to finalise the book. With many authors contributing to the 
book, there is a need to balance authors' autonomy so that the final version of 
the book have a consistent look/design. One mean to achieve this is by producing 
visualizations that meet a set of criteria.

A style guide can be used to create a cohesive edited book. Implementing a 
style guide, however, is time consuming and expensive. The editors/designers
need to spend a lot of effort and time to update or reproduce figures. Another
option that can be taken is by creating an R package that contains theme and 
colour scale functions that work with `ggplot2` [@ggplot2]. This will allow users
(in this case, authors) to add a theme and a colour scale function to their
plots. This is easy to use, and it will cause minimal disruptions to authors'
workflow. More importantly, this will allow the book to have visualizations that are
readable, cohesive, and accessible.

There are several aspects that need to be considered to create a cohesive edited
book. Firstly, a ggplot theme needs to be developed in a way that allows readers
to extract information from the visualizations as much as possible. In addition
to theme, various colour scales that are accessible need to be developed for 
different data types (e.g. categorical and numerical information). The theme
and colour scales will be wrapped in an R package called `dataviz`. The R package
contains the following items:

* Function for the theme and colour scales.
* Documentations.
* Testing.

In the next section, the project will discuss how a theme that allows viewers
to extract information as much as possible from a graph was developed for 
the project. Next, the project will explore the theory behind colour scales in
data visualizations, and describe the methods of choosing colour palettes for
the project. In the fourth section, the report will explain how the theme
and colour schemes will be wrapped in an R package. To put it briefly, the
report contributes themes, colour schemes, and a package called `dataviz`.

# Themes

The first aspect that needs to be considered is how a good visualization looks 
like, and whether there are general rule that can be applied across different 
types of plots. However, before looking at principles of good data visualizations, 
it is better to understand what constitute a visualization. Azzam, Evergreen, and 
Germuth [-@azzam2013data] stated that data visualization is defined by three 
criteria: 

* A process that is based quantitative or qualitative data
* The result of the process is a graph that is representative of the raw data
* The graph is readable and support exploration, examination, and communications
of data.

Without careful consideration, data visualization has the 
potential to cause confusion resulting in misunderstanding or error in decision
making [@evergreen2011eval]. In addition, inefficient visual decoding may result in
users to not detect important aspect of the data [@cleveland1993model]. 
Unfortunately, inefficient graphs are not uncommon occurrence, especially with 
increasing number of options in data visualization over the years. Tufte 
[-@tufte2001visual] found that prevalent errors and data misrepresentation were 
found in mass media. Another study found that a total of 1,365 cases of unnecessary
graphical elements were found out of 20,080 visualizations from a set of articles
published between 1996 and 2016 [@friedman2021data].

With widespread error in data visualizations, guidelines for better graphs
for quantitative and qualitative data were established. According to Tufte 
[-@tufte2001visual], visualizations need to bring viewer's attention to the 
substance of the data as opposed to graphic design, methodology or something 
else. In addition, graphs should prevent distortion of the data and should 
encourage viewers to compare different parts of data. Another suggestion from 
Tufte is to avoid the usage of chartjunk. Any decorations that do not provide 
new information to readers are considered to be chartjunk. This can easily cause 
viewers to be confused or distracted from the data. Some examples of chartjunk 
are moiré effect (the illusion of vibration caused by excessive patterns), 
heavy grid lines, noisy background, a fake perspective (i.e. 3-D charts).

Cleveland and McGill [-@cleveland1984graphical] conducted a study of effective
data visualization based on graphical perception to investigate humans' ability 
to visually decode information encoded on graphs. In the study, people were 
given elementary perceptual tasks to decode information from graphs. The tasks 
are then ordered based on the accuracy of the performance. The order of the 
elementary tasks from most to least accurate is:

* Position along a common scale
* Positions along nonaligned scales
* Length, direction, angle
* Area
* Volume, curvature
* Shading, colour saturation

This shows that there are general rules that can be applied across different
types of plots, and this can be computationally coded once for efficiency.
The next section will discuss how this can be implemented through the use of
`ggplot2` [@ggplot2] theme.

## Roles of Themes in Data Visualization

Themes in `ggplot` [@ggplot2] package are *non-data* components of data visualizations.
Themes allow user to customize plot the background, axis line, ticks, legend 
position, grid, etc. As a default, `ggplot` [@ggplot2] uses `theme_grey` for plots. 
However, `ggplot` [@ggplot2] provides other complete themes that users can use, such as 
`theme_bw()` and `theme_minimal()`.

Theme is especially important for a research book where there are many authors
contributing to the book. Without a set theme, it is likely that each author 
will apply their own visualization styles for their research. Not only does 
setting a theme for a research book allow authors to avoid spending their time 
to customize their plots' appearance, but it also allows research book to 
have consistency in their visualizations.

As stated in previous section, customizing plot themes is imperative in data 
visualizations as non-data components impacts readers' ability to comprehend the 
graph. The next section will discuss how general principles of graphical excellence
from Tufte can be applied to themes.

## Criteria for good themes

There are five principles of data graphics that are proposed by Tufte 
[-@tufte2001visual]. This project focuses on two of the principles that 
specifically relate to a theme. In his book, Tufte introduced an important
concept: data-ink. Data-ink is defined as a non-redundant ink; it is ink used 
on a graph that represent the underlying data. 

$$
\text{Data-ink ratio} = \frac{\text{data-ink}}{\text{total ink used to print the graphic}}
$$

Good graphical representations can be achieved by reducing non-data ink as this
a way to maximize data-ink ratio [@tufte2001visual]. Graphical inks that 
provide no new information about the underlying data are not beneficial to 
readers and sometimes result graphical clutters. However, maximizing the 
data-ink ratio needs to be done cautiously as components of the graph that 
provide useful information can be removed by mistake.

Design choices on data visualizations need to allow information to be presented in an
organised fashion. For example, the colour of non-data components (axis, labels, 
grid lines, borders) can be grey, and the data component can be brightly
coloured, making the data to be the focus of attention [@stone2006choosing]. 
This will assist readers to better understand the roles and relationship between 
different elements of the visualizations.

## Potential Themes

```{r scatter-grey, fig.cap="A scatter plot example from `mtcars` dataset with `theme_grey()`. This theme uses light grey background with white grid lines.", fig.height=5, fig.width=5}
g1 <- ggplot(mtcars, aes(mpg,disp,color=factor(carb))) + 
  geom_point(size=3) + 
  labs(title="Scatter Plot",
       subtitle = "theme_grey()",
       caption = "Data: mtcars")

g1
```

The defualt `ggplot2` theme is `theme_grey()`, with grey background and white 
grid lines. These colour combinations reduces the contrast between the background 
and grid lines and gives the data plots more value-added appearance 
[@carr1994using]. Moreover, grey background allows the plot to have a similar
colour to the rest of the texts. As a result, the graphics blend well with
the flow of a text [@WickhamHadleygEGf].

While having a grey background and white grid
lines have their benefits, some steps can be made to improve the theme. As
shown in figure \@ref(fig:scatter-grey), having a grey panel background  
on top of a white plot background unnecessarily increase non-data components.
In addition, having both major and minor grid lines distracts readers from the
real data, and rarely needed as data visualizations are not meant to provide
precise quantitative value [@few2005grid].

A ggplot theme is developed for this project that is not only based on 
Tufte principles of data graphics, but also to create a theme that is suitable 
for a wide range of plots. There are two potential themes that were developed for 
`dataviz`: `theme_seagull()` and `theme_wombat()`.

### Seagull Theme

```{r theme-comparison, fig.cap="Comparison between `theme_seagull` (left), `theme_wombat` (middle), and `theme_bw` (right). The rows represent different figure types from a data example from R. Basic colour palette from R is used to show categorical variables for this example. In the next section, we will discuss accessible colour palettes that are developed for the project.", fig.height=13, fig.width=11}
w1 <- ggplot(mtcars, aes(mpg,disp,color=factor(carb))) + 
  geom_point(size=3) + 
  labs(title="Scatter Plot",
       subtitle = "theme_wombat()",
       caption = "Data: mtcars") +
  theme_wombat()

w2 <- ggplot(mtcars, aes(factor(carb),fill=factor(carb))) + 
  geom_bar() + 
  scale_y_continuous(expand = c(0,0)) + 
  scale_x_discrete(expand = c(0,0)) +
  labs(title="Bar Plot",
       subtitle = "theme_wombat()",
       caption = "Data: mtcars") +
  theme_wombat()

w3 <- ggplot(economics_long, aes(date, value01, colour = variable)) +
  geom_line() +
   scale_y_continuous(expand = c(0,0)) + 
   scale_x_date(expand = c(0,0)) +
  labs(title = "Line Plot",
       subtitle = "theme_wombat()",
       caption = "Data: economics_long") +
  theme_wombat()

s1 <- ggplot(mtcars, aes(mpg,disp,color=factor(carb))) + 
  geom_point(size=3) + 
  labs(title="Scatter Plot",
       subtitle = "theme_seagull()",
       caption = "Data: mtcars") +
  theme_seagull()

s2 <- ggplot(mtcars, aes(factor(carb),fill=factor(carb))) + 
  geom_bar() + 
  scale_y_continuous(expand = c(0,0)) + 
  scale_x_discrete(expand = c(0,0)) +
  labs(title="Bar Plot",
       subtitle = "theme_seagull()",
       caption = "Data: mtcars") +
  theme_seagull()

s3 <- ggplot(economics_long, aes(date, value01, colour = variable)) +
  geom_line() +
  scale_y_continuous(expand = c(0,0)) + 
  scale_x_date(expand = c(0,0)) +
  labs(title = "Line Plot",
       subtitle = "theme_seagull()",
       caption = "Data: economics_long") +
  theme_seagull()

bw1 <- ggplot(mtcars, aes(mpg,disp,color=factor(carb))) + 
  geom_point(size=3) + 
  labs(title="Scatter Plot",
       subtitle = "theme_bw()",
       caption = "Data: mtcars") +
  theme_bw()

bw2 <- ggplot(mtcars, aes(factor(carb),fill=factor(carb))) + 
  geom_bar() + 
  labs(title="Bar Plot",
       subtitle = "theme_bw()",
       caption = "Data: mtcars") +
  theme_bw()

bw3 <- ggplot(economics_long, aes(date, value01, colour = variable)) +
  geom_line() +
  labs(title = "Line Plot",
       subtitle = "theme_bw()",
       caption = "Data: economics_long") +
  theme_bw()

(s1 + w1 + bw1) / (s2 + w2 + bw2) / (s3 + w3 + bw3)
```

Plot examples with `theme_seagull()` can be seen in figure \@ref(fig:theme-comparison). 
This `ggplot2` theme is built based on `theme_minimal()`. The idea behind this
theme is to reduce non-data components as much as possible by maximizing white 
space around the plot. The `%+replace%` operator is used to change the elements of 
the plots. Axis line on the y axis is removed, and axis line on the x axis is 
thicker. The major vertical grid lines and minor grid lines on both axis are removed. 
The grid lines, as shown in figure \@ref(fig:theme-comparison), are visually 
muted so that it does not distract the data points. Cleveland 
[-@cleveland1993visualizing] found that grid lines enhance readers' perceptions 
in making visual comparison and accuracy of information extraction. 

### Wombat Theme

Another potential theme for this project is called `theme_wombat()`.
This theme is developed  based on `ggplot2` basic theme `theme_bw()`. 
The `%+replace%` operator is also used to develop `theme_wombat()`. It can be seen
in figure \@ref(fig:theme-comparison) that minor grid lines on both axis are 
removed for this theme. Similar to `theme_seagull()`, `theme_wombat()` has 
white background and light grey grid lines.

## Themes Comparison

This section is going to compare the potential themes `theme_seagull()` and
`theme_wombat()` with `ggplot2` [@ggplot2] built in theme, `theme_bw()`. 
The `theme_bw` is chosen for this comparison as it is one of the built-in themes 
that minimizes non-data components while still preserving light coloured grid lines.

Side by side comparison of scatter plots, bar charts, and line charts between
the three themes can be found in figure \@ref(fig:theme-comparison). One of 
the differences between `dataviz` potential themes and `theme_bw()` is the size
of the text. The axis title, plot title, subtitle, and caption with 
`theme_seagull()` and `theme_wombat()` are easier to read in comparison to
`theme_bw()`. Secondly, the legend position is located on the bottom of the
plot as opposed to the right of the plot. The reason to the change in legend
position is to maximize the data-ink ratio of the plots. Putting the legend
on the right of the plot takes up the space that can be used to display the
underlying data. Another way of maximizing data-ink ratio is by reducing
the margin around the plot. As shown in the bar plots and line plots in figure \@ref(fig:theme-comparison), there are no extra space between the bars/lines 
with the axis lines in the plots with `theme_seagull()` and `theme_wombat()`.

One of the objectives of `theme_seagull()` is to have a theme that increases 
data-ink ratio as much as possible. In comparison to other two themes, 
`theme_seagull()` has the least amount of grid lines and borders around the
plot, making it to have the highest data-ink ratio. While there is no vertical 
grid lines, major horizontal grid lines are preserved as they aid readers to 
approximate and compare values between data points.

The `theme_wombat()` is developed to be the middle ground between `theme_seagull()`
and `theme_bw()`. In comparison to the `ggplot2` built-in theme, this theme
is less cluttered as it does not have minor grid lines on both axis. Unlike
`theme_seagull()`, however, this theme retains major vertical grid lines.

### The Chosen Theme

As stated above, `theme_seagull()` has the least clutter between the two 
potential themes.However, `theme_wombat()` is likely to assist readers
to perform comparison more accurately than `theme_seagull()`
while still maintaining a high level of data-ink ratio. Using the scatter plot 
example in figure \@ref(fig:theme-comparison), the major vertical grid lines 
from `theme_wombat()` aids readers' ability to estimate the value of data points along the
x axis. This shows that `theme_wombat()` is a good compromise of minimizing 
clutters as well as allowing readers to extract information from the graph as 
much as possible.

# Choosing Colour Palletes

In addition to themes, colour also plays an important role in data 
visualizations. Proper use of colour in a visualization enhances the quality
of information being presented [@zeileis2009escaping]. One of the most important
use of colour in data visualizations is to differentiate one element from one
another [@stone2006choosing]. However, a poor choice of colour palette can impede 
viewers' ability to effectively extract information from the visualization. 

With a simple visual features, some objects can be easily recognised compared
to other objects without requiring attention. The target object will "pop-out" 
in field of vision regardless how many distractors surrounding it 
[@treisman1985preattentive]. The use of colour in data visualization can facilitate
this "pop-out" as it makes some elements to be easier to be found.

```{r pop-out, fig.cap="This plot shows the same data with groups highlighted with either colour (left hand panel) or shape (right hand panel). From this, we see that it is easier to find the odd ones by using colour.", fig.height=3}
set.seed(20211028)

x <- rnorm(20)
y <- rnorm(20)
color <- c("red","red","blue","red","blue","red","red","red","red","red","red",
           "blue","red","red","red","red","red","red","red","red")
pop_out <- data.frame(x,y,color)

p1 <- pop_out %>% ggplot(aes(x = x,
                             y = y,
                             color = color)) +
  geom_point() +
  theme_wombat() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  ggtitle("Color")

p2 <- pop_out %>% ggplot(aes(x = x,
                             y = y,
                             shape = color)) +
  geom_point() +
  theme_wombat() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  ggtitle("Shape")

p1 + p2
```

A great example to show how easy it is for colour to be recognised is stated
in Healy's book [@healy2018data]. Figure \@ref(fig:pop-out) contains three "odd" 
observations that are different 
than the rest. To represent this, the first panel uses colour to differentiate 
the different kinds of observations. The three observations are coloured red, and
the rest are blue. On the other hand, the second panel uses shape to represent
this feature. Circle is used to represent the "odd" observations, and triangle is
used for the rest. While readers have no problem in identifying the three 
observations in the second panel, it is much easier to find these observations 
on the first panel. This shows that finding the observations of interest in the 
first panel requires significantly lower level of attentions compared to the 
second panel.

In addition to assisting viewers in differentiating one attribute to another, 
colours can also be used to evoke viewers' emotions. Emotion is important
in visualizations for story telling, communicative intent, and engagement
[@bartram2017affective].  It was also found that emotions can significantly 
improve many cognitive processes such as attention, learning, decision making, 
and problem-solving [@harrison2012exploring].

This project will focus to create colour schemes that assist users to easily
differentiate one data attribute from another. To achieve this, it is important
to choose these schemes using a colour space that works well with human perception
of colour. The next section will discuss the different options of colour space,
and which one will be used for the project.

## Colour Spaces

To effectively use colours in visualizations, it is imperative to understand
how human colour perception works. It was theorized that human colour vision evolved 
in three distinct stages [@ihaka2003colour]: 

1. Perception of light and dark (monochrome)
2. Perception of yellow/blue contrast
3. Perception of green/red contrast

As a result of these three colour axes, colours are naturally described as 
locations in a three dimensional space. While these colour axis provide 
physiological description, human perception of colours do not corresponds to
these axes. Human perception of colour corresponds to the polar coordinates
of the colour coordinate (yellow/blue and green/red), and a dark/light axis 
[@zeileis2006choosing].

Unfortunately, not all software packages use colour space was developed based on
human perception. For example, the most common colour map used by software 
packages, the Red-Green-Blue (RGB) [@stauffer2015somewhere], was primarily 
developed to correspond with colour generation on computer screens rather than 
to correspond with human perception of colour [@zeileis2009escaping]. 
Zeileis, Hornik, and Murrell further stated that it is practically impossible
for a human to control the perceptual properties of a colour within this colour 
model as there is no single dimension that relates to hue or brightness of a 
colour

One popular alternative to RGB system is Hue-Saturation-Value (HSV). This model
is a simple transformation of RGB system and aims to capture the dimensions of 
human visual perceptions [@smith1978color]. Unfortunately, the result of mapping 
HSV's three dimensions to three dimensions of human perceptions is very poor
[@zeileis2006choosing]. Moreover, it is fairly difficult to find a set of colours 
that are in harmony in HSV colour space [@zeileis2006choosing]. Furthermore, HSV encourages 
the use of colours that are highly saturated. While these colours are good to 
attract viewers' attention, it is difficult to look at for a long time 
[@zeileis2009escaping].

HCL, which is short for Hue-Chroma-Luminance, was developed to overcome many 
problems encountered by other colour spaces. This colour space is developed
based on the three colour attributes that corresponds to human perceptions:
hue, chroma, and luminance [@sarifuddin2005new]. It was found that the three
dimensions from this colour space matches human visual perception well 
[@zeileis2019colorspace]. In general, HCL colour space can be used to generate 
three types of colour scales: qualitative, sequential, and diverging.

### Qualitative Palettes

This colour palette is designed for coding categorical information where the 
categories do not have particular ordering [@zeileis2019colorspace]. In other
words, this colour palette will give the same perceptual weight to each category
so that there will be no perception of one group to be larger or more important
than any other one [@zeileis2006choosing]. This colour palette is commonly 
applied to bar plots or box plots.

### Sequential Palettes

Sequential pallets are used for coding ordered/numerical information that span
in a certain interval [@zeileis2006choosing]. The low values of the interval
are considered to be uninteresting, whereas high values are considered to be 
interesting [@zeileis2019colorspace]. Low values are usually depicted by lighter
colours, while darker colours are used for high values [@harrower2003colorbrewer].
This palette is usually used on heatmaps.

### Diverging Palletes

Similar to sequential palette, diverging pallets are designed for coding ordered/
numerical information that ranges in an interval. The difference between 
diverging and sequential palletes is that diverging pallets include a neutral
value [@zeileis2006choosing]. This palette should be used when an important data
class or break points need to be highlighted [@harrower2003colorbrewer]. Harrower 
and Brewer further stated that this critical data class or break points is 
highlighted by a change in in hue and luminance, and should represent a critical 
value of the data (e.g. mean, median, or zero). For instance, the average GDP
per-capita in a choropleth map can be emphasized (i.e. set as neutral value) so 
that the values above and below the average can be shown in different hues.

## Colour palette of choice

In addition to plot themes, this project also contributes two colour schemes for 
each of the three colour palettes type mentioned in the previous section. 
The `pals` packages used to build the colour palettes 
and assess them. The chosen colour palettes need to be assessed to ensure
that they achieve their intended purpose. For example, the qualitative colour palette
needs to avoid implicit ordering as this palette is intended for representing categorical 
variables. In addition, we need to ensure that people with colour vision
impairments are able to distinguish the different colours in the chosen palettes.
The `pal.safe()` function from `pals` package allows us to see how colour palettes
are seen by deutan, protan, and tritan colour-blind people.

The methodology of choosing colour palettes that are not only accessible but also 
enhance the quality of visualizations will be discussed in this section. Firstly, the
report will explore common strategies used in choosing qualitative, sequential, 
and diverging palettes. Then, examples of palettes made from these strategies will
be assessed. Lastly, the report will show how colour schemes for 
qualitative, sequential and diverging palettes are chosen for the project.

### Chosen Qualitative Palettes

```{r color-blind, fig.height=8, fig.cap="Colour palettes as seen by people with colour vision impairment. The first six colour palettes the palettes that were developed for this project. The last palette, Harmonic palette, is a colour palette from `pals` package."}
quokka_palette <- c("#8C01E6", "#93B1EB", "#AB313D", "#D6CA24")
quoll_palette <- c('#8D5698', '#EFD36F', '#247A3E', '#17B0A8')
seal_palette <- colorspace::sequential_hcl(7,
                                              h = c(0, -140),
                                              c = c(80,40),
                                              l = c(40, 75),
                                              power = c(1,1))
snake_palette <- colorspace::sequential_hcl(6,
                                              h = c(250, 90),
                                              c = c(40, 55),
                                              l = c(33,98),
                                              power = c(0.5,1),
                                              rev = TRUE)
dingo_palette <- colorspace::diverge_hcl(7,
                                           h = c(260, 0),
                                           c = 80,
                                           l = c(30, 90),
                                           power = 1.5)
dolphin_palette <- colorspace::diverging_hcl(n = 7,
                                             h = c(265, 45),
                                             c = 80,
                                             l = c(70, 95),
                                             power = c(0.5, 2),
                                             rev = TRUE)

# colorblind filter
par(mfrow=c(4,2))
pal.safe(quokka_palette, main = "Quokka Palette")
pal.safe(quoll_palette, main = "Quoll Palette")
pal.safe(seal_palette, main = "Seal Palette")
pal.safe(snake_palette, main = "Snake Palette")
pal.safe(dingo_palette, main = "Dingo Palette")
pal.safe(dolphin_palette, main = "Dolphin Palette")
pal.safe(qualitative_hcl(4, palette = "harmonic"), main = "Harmonic Palette")
```
\newpage

```{r harmonic, fig.cap="HCL colour spectrum plot for Harmonic palette. This palette has constant luminance and chroma, but varying hues across the four colours."}
specplot(qualitative_hcl(4, palette = "harmonic"), main = "Harmonic Palette")
```

One common strategy to choose this type of palette is by keeping chroma and luminance
relatively constant, and hue varied; making all colours in the palette to balance 
towards the same grey [@zeileis2009escaping]. While this strategy is relatively
simple, keeping chroma and luminance fixed may not be the best approach to 
choose colour palette that are accessible.

Figure \@ref(fig:harmonic) shows an example of a qualitative colour palette from 
`colorspace` package called harmonic. As shown in the figure, chroma
and luminance remain constant, while the hue varies across the four colours.
It is fair to say that this colour set works relatively well to show categorical 
information as it does not give implied ordering across different categories.
This method, however, does not have the same result for people with visual
impairment. It can be seen in figure \@ref(fig:color-blind) that people 
with colour-blindness will have difficulty in distinguishing some colours from 
harmonic palette. For instance, there are three colours that look similar as seen 
by people with tritan colour-blindness. Moreover, this colour palette results in 
implied ordering of different data group for people with visual impairment.

```{r qualitative, fig.height=3, fig.cap="HCL colour spectrum plot for Quokka and Quoll palette. The hues, chroma, and luminance vary across the four colours for Quokka and Quoll palettes. These varying hues, chroma and luminance avoid natural ordering."}
par(mfrow=c(1,2))
specplot(quokka_palette, main = "Quokka Palette")
specplot(quoll_palette, main = "Quoll Palette")
```

The two qualitative colour palettes developed for the project are called 
Quokka and Quoll palettes. Figure \@ref(fig:qualitative) shows 
the HCL colour spectrum plot for these palettes. In
contrast with the Harmonic Palette, Quokka and Quoll palettes have differing 
hues, chroma and luminance across all colours. As a result, each colour in
these palettes can easily be distinguished by people with deutan, protan, and 
tritan colour-blindness (see figure \@ref(fig:color-blind)). Furthermore,
natural ordering implied by Quokka and Quoll is less prominent than Harmonic
palette.

### Chosen Sequential Palettes

The strategy to choose a sequential palette is different to that of qualitative
palette. One easy way to create a sequential palette is by choosing two values 
for hue (h1 and h2), chroma (c1 and c2), luminance (l1 and l2), and power 
transformations (p1 and p2) for chroma and luminance [@zeileis2006choosing].
These values of hues, chroma, luminance, and power transformation are used
as inputs to `sequential_hcl` function to produce a set of sequential palette.
The power transformation allows the chroma and luminance to increase from low
values to high values in a nonlinear fashion.

```{r sequential, fig.height=3, fig.cap="HCL colour spectrum plot for Seal and Snake palettes. These palettes range between two hues from darker to lighter shades (or vice versa)."}
par(mfrow=c(1,2))
specplot(seal_palette, main = "Seal Palette")
specplot(snake_palette, main = "Snake Palette")
```

The two sequential palettes created for this project are Seal and
Snake palettes. As shown in figure \@ref(fig:sequential),
these palettes use multi-hue schemes as opposed to single-hue schemes. Multi-hue
palettes were chosen for the project as it provides better contrast between different
values compared to a single-hue sequential palette [@harrower2003colorbrewer].
Moreover, multi-hue sequential palettes avoid the use of extreme colours, as well
as allowing viewers to better discriminate the middle colours [@zeileis2019colorspace]. 
In addition, figure \@ref(fig:color-blind) shows that people with deutan, protan, 
and tritan colour blindness are able to distinguish the colours in the palette.
This means that the colour palettes will allow readers with colour vision impairment
to see implied order and how the data ranges from low to high values.

### Chosen Diverging Palettes

The approach in choosing diverging palettes is similar to choosing sequential
palettes. Diverging palettes can be built by choosing two values for
hues (h1 and h2), chroma (c1 and c2), luminance (l1 and l2), and power 
transformation (p1 and p2) for chroma and luminance [@zeileis2006choosing]. 
These values will be taken as inputs to function `diverge_hcl()` to create a diverging 
palette. It will automatically choose the neutral colour for the palette.

```{r diverging, fig.height=3, fig.cap="HCL colour spectrum plot for Dingo and Dolphin palette. Similar to sequential palettes, these palettes range between two hues from darker to lighter shades. The only difference is that this palette has a neutral colour."}
par(mfrow=c(1,2))
specplot(dingo_palette, main = "Dingo Palette")
specplot(dolphin_palette, main = "Dolphin Palette")
```

Three conditions need to be met for for a diverging palette to avoid colour 
confusion. Colour confusion are caused by naming confusions, colour vision 
impairment, and simultaneous contrast confusion [@brewer1996guidelines]. 
Naming confusion happens when people use two basic colour names for the same 
colour (e.g. a colour was called red when given once, then pink when given 
another time). On the other hand, simultaneous contrast
refers to a phenomena where colours are affected by their opposite (e.g.
a colour will appear lighter when surrounded by a dark colour, and a colour will
appear warmer when surrounded by a cool colour) [@brewer1992review]. This can be
prevented by avoiding selecting pairs of colours that are opposite in the colour 
wheel for diverging palettes [@brewer1996guidelines].

Dingo and Dolphin palettes are diverging palettes developed for the project.
These palettes are shown in figure \@ref(fig:diverging).
Red/blue pairs and orange/blue hue pairs are chosen for Dingo and Dolphin
palettes respectively as these pairs create no naming confusions and 
simultaneous contrast confusions [@brewer1996guidelines]. Moreover, these 
palettes create no colour confusion due to colour vision impairment as colours in 
Dingo and Dolphin palettes can be easily distinguished by people with colour-blindness
(shown in figure \@ref(fig:color-blind)). Another reason of choosing these pairs
is because red/blue pair is often used in the topic of the edited book, MRP, 
to show electoral result (e.g. Democratic and Republican party). Therefore,
it is important to have a red/blue scale and a non-red/blue scale for functionality
in the book.

# R Package

One of the objectives of this project is to create a solution for authors and
editors of the research book to create visualizations that are readable, 
cohesive, and accessible. The solution needs to be easy to use so that it
does not disrupt authors' workflow. To achieve this, R code necessary
to create the chosen `ggplot2` [@ggplot2] theme and colour schemes will be bundled in an R 
package called `dataviz`. R packages are extensions to R that can comprise of 
codes, data, documentation, and tests [@wickham2015r]. 

One of the reasons for creating an R package for this project is because of the
ease of use. Codes, documentation, and tests in an R package are built in a 
standardized format so that any R user can install it [@wickham2015r]. In 
addition, it is fairly simple to create an R package. Packages such as `roxygen2` 
[@roxygen2], `devtools` [@devtools], and `testthat` [@testthat] provide great 
assistance in documenting and testing the package. Another benefit of using an 
R package is that any updates to the theme and colour schemes are not time 
consuming for the users. Any updates to themes and colour schemes simply need to 
be applied to the code in the R package. There is no need to update the code 
for each plots in the book; it just needs to be recompiled. This shows that 
using an R package will allow us to achieve one of the aims of the project: 
to provide a solution that causes minimal disruption to authors' work.

For users to be able to easily access an R package, it needs to be stored in a 
repository. For this project, the `dataviz` package will be stored in GitHub.
GitHub is a popular choice for a repository as it has the integration with git, 
a version control software. Version control records changes applied to the 
package to a file. This allows user to recall specific versions of the package
at later time [@loeliger2012version]. An alternate option to GitHub is 
Comprehensive R Archive Network (CRAN), however, this is not suitable for this 
package as `dataviz` has a fairly small scope and is intended for a small group 
of people (i.e. authors of the MRP research book). CRAN is usually a repository 
used for more complex R packages intended for large group of people.

In the next section, the report will discuss how the `dataviz` package was 
created. It will discuss each component of the package, and what tools 
were used in developing the package.

## `dataviz` Components

One of the most important components of a package is the R code. One firm
rule for an R package is that its R functions need to be stored in R scripts 
[@wickham2015r]. Following this rule, the `dataviz` has its R code in  R 
scripts, which are stored in `R` directory. The chosen colour palettes can be 
accessed from the `scale_colour_dataviz` and `scale_fill_dataviz` functions which
are stored in `colour.R`. These functions take the name of the palette as an 
input (e.g. quokka). The ggplot theme chosen for the project can be used by
function `theme_wombat()` which is stored in an R script called `theme.R`. 

In addition to R codes, documentation is also a crucial aspect of a well-made
package [@wickham2015r]. In the absence of documentation, users will not have a guide
that assists them to use the package properly. Proper documentation also assists
the authors of the package for future use. This assists authors to remember the
purpose of the functions. One type of documentation is an object documentation;
this can be accessed by the `?` operator or `help()`. Information such as function 
description, usage, argument(s), and examples can be found in the documentation.
The `roxygen2` [@roxigen2], `devtools` [@devtools], and `usethis` [@usethis] 
packages are used in creating object documentation.

Another crucial aspect to package development is testing. While testing add
extra work to package development, it ensures that the package works
as intended. It also ensures that the package works for "edge cases" - the less
common uses that a user might use. In addition, testing ensures error and warning
messages to appear when required (and not when they are not supposed to appear).
In other words, testing provide protection and quality assurance for users.
We use the packages `testthat` [@testthat] and `usethis` [@usethis] to write 
the tests for the `dataviz` package. One example of a test that was used for 
this package is one that ensures that users use valid input when using 
`scale_colour_dataviz` and `scale_fill_dataviz`.

\newpage

# Conclusions and Limitations

One of the challenges in producing a research book is creating a book that has
a consistent look or design. Typically, there is a group of authors that work on
every chapter of an edited book. These authors will have their own unique style in
producing their visualizations for the book. While this is not a bad thing, 
visualizations style needs to be standardized to keep the look of the book 
consistent. Not only do these visualizations need to have a consistent design, 
but they also need to be readable and accessible.

The aim of the project is to provide a solution that can ensure that 
visualizations in the research book meet the criteria mentioned in the previous
paragraph in an efficient manner. To achieve this objective, a ggplot theme was 
developed for the project. The chosen theme was developed based on Tufte's
principles of graphical excellence. In addition, several colour scales for various
data types were also created. The colour palettes were developed using a colour 
space that was based on human perception. Furthermore, to ensure that color is 
accessible, we visualized how these colours as they would be seen by people with 
colour visual impairment. These theme and colour scales were developed to ensure
that the viewers to extract information as much as possible from the 
visualizations. The theme and colour schemes are bundled in an R package for 
efficiency.

One limitation in the package is related to colour palettes, specifically
the chosen qualitative palettes. As mentioned in the previous section,
this palette is specifically used to show *un-ordered* categorical information.
It is therefore imperative to avoid implying that one group is larger or more
important than other groups. However, removing implied ordering while ensuring 
accessibility proved to be difficult. Figure \@ref(fig:color-blind) shows that
while the Quokka and Quoll palettes are better than Harmonic palette in
avoiding implied ordering for people with visual impairment, there are still 
some colours that look paired. In addition, Quokka and Quoll only have four
colours. It is difficult to produce a qualitative palette that have more than 
four colours while still ensuring that people with color impairment are able to 
distinguish the colours from one another.

One area of future work for this project is to work on writing additional 
testings. A test can be developed to ensure that a warning message will appear
when the theme function is applied to a non-ggplot object. Another test that 
can be applied to the package is to ensure an error to pop-up when applying
a diverging palette to a categorical variable. Another approach that can be 
done in the future is by having the users (i.e. the authors of the MRP books)
to test the functionality of the package for a period of time. This allows
the users to provide feedback, and this can be used to improve the package.
These testings are beneficial to protect the quality of the package.

In summary, this project contributes an R package containing a ggplot theme and
six colour palettes developed to assist MRP book authors in creating visualizations that
are readable, cohesive, and accessible.

\newpage

# Reference
