---
title: "Project Report"
author: "Jimmy Effendy"
date: "13/10/2021"
output: bookdown::pdf_document2
bibliography: references.bib
csl: apa.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(ggplot2)
library(tidyverse)
library(dataviz)
library(colorspace)
library(patchwork)
library(pals)
```

\newpage

# Introduction

This project relates to an edited book about multilevel regression and 
poststratification (MRP). As is the case with other books, there will be
numerous visualizations that will be used within the book. These plots will need to be
readable, cohesive, and accessible. There is a group of authors (around one to 
five authors) who work together on a chapter. The editors then work to combine 
all the chapters to finalise the book. With many authors contributing to the 
book, there is a need to balance authors' autonomy so that the final version of 
the book have a consistent look/design. One mean to achieve this is by producing 
visualizations that meet a set of criteria.

A style guide can be used to create a cohesive edited book. Implementing a 
style guide, however, is time consuming and expensive. The editors/designers
need to spend a lot of effort and time to update or reproduce figures. Another
option that can be taken is by creating an R package that contains theme and 
color scale functions that work with `ggplot2`. This will allow users
(in this case, authors) to add a theme and a color scale function to their
plots. This is easy to use, and it will cause minimal disruptions to authors'
workflow. More importantly, this will allow the book to have visualizations that are
readable, cohesive, and accessible.

There are several aspects that need to be considered to create a cohesive edited
book. Firstly, a ggplot theme needs to be developed in a way that allows readers
to extract information from the visualizations as much as possible. In addition
to theme, various color scales that are accessible need to be developed for 
different data types (e.g. categorical and numerical information). The theme
and color scales will be wrapped in an R package called `dataviz`. The R package
contains the following items:

* Function for the theme and color scales.
* Documentations.
* Testing.

In the next section, the project will discuss how a theme that allows viewers
to extract information as much as possible from a graph was developed for 
the project. Next, the project will explore the theory behind color scales in
data visualizations, and describe the methods of choosing color palettes for
the project. In the fourth section, the report will explain how the theme
and color schemes will be wrapped in an R package. To put it briefly, the
report contributes themes, color schemes, and a package called `dataviz`.

# Themes

The first aspect that needs to be considered is how a good visualization looks 
like, and whether there are general rule that can be applied across different 
types of plots. However, before looking at principles of good data visualizations, 
it is better to understand what constitute a visualization. Azzam, Evergree, and 
Germuth [-@azzam2013data] stated that data visualization is defined by three 
criteria: 

* A process that is based quantitative or qualitative data
* The result of the process is a graph that is representative of the raw data
* The graph is readable and support exploration, examination, and communications
of data.

Without careful consideration, data visualization has the 
potential to cause confusion resulting in misunderstanding or error in decision
making [@evergreen2011]. In addition, inefficient visual decoding may result in
users to not detect important aspect of the data [@cleveland1993model]. 
Unfortunately, inefficient graphs are not uncommon occurrence, especially with 
increasing number of options in data visualization over the years. Tufte 
[-@tufte2001visual] found that prevalent errors and data misrepresentation were 
found in mass media. Another study found that 30% of graphs found in Volume 207 
(1980) of *Science* have one or more type of errors [@cleveland1984graphs].

With widespread error in data visualizations, guidelines for better graphs
for quantitative and qualitative data were established. According to Tufte 
[-@tufte2001visual], visualizations need to bring viewer's attention to the 
substance of the data as opposed to graphic design, methodology or something 
else. In addition, graphs should prevent distortion of the data and should 
encourage viewers to compare different parts of data. Another suggestion from 
Tufte is to avoid the usage of chartjunk. Any decorations that do not provide 
new information to readers are considered to be chartjunk. This can easily cause 
viewers to be confused or distracted from the data. Some examples of chartjunk 
are moiré effect, heavy grid lines, noisy background, a fake perspective 
(i.e. 3-D charts).

Cleveland and McGill [-@cleveland1984graphical] conducted a study of effective
data visualization based on graphical perception to investigate humans' ability 
to visually decode information encoded on graphs. In the study, people were 
given elementary perceptual tasks to decode information from graphs. The tasks 
are then ordered based on the accuracy of the performance. The order of the 
elementary tasks from most to least accurate is:

* Position along a common scale
* Positions along nonaligned scales
* Length, direction, angle
* Area
* Volume, curvature
* Shading, color saturation

This shows that there are general rules that can be applied across different
types of plots, and this can be computationally coded once for efficiency.
The next section will discuss how this can be implemented through the use of
`ggplot2` theme.

## Roles of Themes in Data Visualization

Themes in `ggplot` package are **non-data** components of data visualizations.
Themes allow user to customize plot the background, axis line, ticks, legend 
position, grid, etc. As a default, `ggplot` uses `theme_grey` for plots. 
However, `ggplot` provides other complete themes that users can use, such as 
`theme_bw()` and `theme_minimal()`.

Theme is especially important for a research book where there are many authors
contributing to the book. Without a set theme, it is likely that each author 
will apply their own visualization styles for their research. Not only does 
setting a theme for a research book allow authors to avoid spending their time 
to customize their plots' appearance, but it also allows research book to 
have consistency in their visualizations.

As stated in previous section, customizing plot themes is imperative in data 
visualizations as non-data components impacts readers' ability to comprehend the 
graph.

Give plot examples.

## Criteria for good themes

There are five principles of data graphics that are proposed by Tufte 
[-@tufte2001visual]. This project focuses on two of the principles that 
specifically relate to a theme. In his book, Tufte introduced an important
concept: data-ink. Data-ink is defined as a non-redundant ink; it is ink used 
on a graph that represent the underlying data. 

$$
\text{Data-ink ratio} = \frac{\text{data-ink}}{\text{total ink used to print the graphic}}
$$

Good graphical representations can be achieved by reducing non-data ink as this
a way to maximize data-ink ratio [@[-@tufte2001visual]. Graphical inks that 
provide no new information about the underlying data are not beneficial to 
readers and sometimes result graphical clutters. However, maximizing the 
data-ink ratio needs to be done cautiously as components of the graph that 
provide useful information can be removed by mistake.

Design choices on data visualizations need to allow information to be presented in an
organised fashion. For example, the color of non-data components (axis, labels, 
grid lines, borders) can be grey, and the data component can be brightly
colored, making the data to be the focus of attention [@stone2006choosing]. 
This will assist readers to better understand the roles and relationship between 
different elements of the visualizations.

## Potential Themes

```{r scatter-grey, fig.cap="A scatter plot example with 'theme_grey()'", fig.height=5, fig.width=5}
g1 <- ggplot(mtcars, aes(mpg,disp,color=factor(carb))) + 
  geom_point(size=3) + 
  labs(title="Scatter Plot",
       subtitle = "theme_grey()",
       caption = "Data: mtcars")

g1
```

`theme_grey()` is the default `ggplot2` theme with grey background and white 
grid lines. These color combinations reduces the contrast between the background 
and grid lines and gives the data plots more value-added appearance 
[@carr1994using]. While having a grey background and white grid
lines have their benefits, some steps can be made to improve the theme. As
shown in figure \@ref(fig:scatter-grey), having a grey panel background  
on top of a white plot background unnecessarily increase non-data components.
In addition, having both major and minor grid lines distracts readers from the
real data, and rarely needed as data visualizations are not meant to provide
precise quantitative value [@few2005grid].

A ggplot theme is developed in `dataviz` package that is not only based on 
Tufte principles of data graphics, but also to create a theme that is suitable 
for a wide range of plots. There are two potential themes that were developed for 
`dataviz`: `theme_seagull()` and `theme_wombat()`.

### theme_seagull

```{r seagull-scatter, fig.cap="A scatter plot example with 'theme_seagull()'", fig.height=5, fig.width=5}
s1 <- ggplot(mtcars, aes(mpg,disp,color=factor(carb))) + 
  geom_point(size=3) + 
  labs(title="Scatter Plot",
       subtitle = "theme_seagull()",
       caption = "Data: mtcars") +
  theme_seagull()

s1
```

This `ggplot2` theme is built based on `theme_minmal()`. The idea behind this
theme is to reduce non-data components as much as possible by maximizing white 
space around the plot. `%+replace%` operator is used to change the elements of 
the plots. Axis line on the y axis is removed, and axis line on the x axis is 
thicker. The major vertical grid lines and minor grid lines on both axis are removed. 
The grid lines, as shown in figure \@ref(fig:seagull-scatter), are visually 
muted so that it does not distract the data points. Cleveland 
[-@cleveland1993visualizing] found that grid lines enhance readers' perceptions 
in making visual comparison and accuracy of information extraction. 

### theme_wombat

```{r wombat-scatter, fig.cap="A scatter plot example with 'theme_wombat()'", fig.height=5, fig.width=5}
w1 <- ggplot(mtcars, aes(mpg,disp,color=factor(carb))) + 
  geom_point(size=3) + 
  labs(title="Scatter Plot",
       subtitle = "theme_wombat()",
       caption = "Data: mtcars") +
  theme_wombat()

w1
```

Another potential theme for `dataviz` package is called `theme_wombat()`.
This theme is developed  based on `ggplot2` basic theme `theme_bw()`. 
`%+replace%` operator is also used to develop `theme_wombat()`. Minor grid 
lines on both axis are removed for this theme. Similar to `theme_seagull()`, 
`theme_wombat()` has white background and light grey grid lines.

## Themes Comparison

This section is going to compare the potential themes `theme_seagull()` and
`theme_wombat()` with `ggplot2` built in theme, `theme_bw()`. `theme_bw` is
chosen for this comparison as it is one of the built-in themes that minimizes
non-data components while still preserving light colored grid lines.

```{r theme-comparison, fig.cap="Comparison between theme_seagull (left), theme_wombat (middle), and theme_bw (right).", fig.height=14, fig.width=11}
w2 <- ggplot(mtcars, aes(factor(carb),fill=factor(carb))) + 
  geom_bar() + 
  scale_y_continuous(expand = c(0,0)) + 
  scale_x_discrete(expand = c(0,0)) +
  labs(title="Bar Plot",
       subtitle = "theme_wombat()",
       caption = "Data: mtcars") +
  theme_wombat()

w3 <- ggplot(economics_long, aes(date, value01, colour = variable)) +
  geom_line() +
   scale_y_continuous(expand = c(0,0)) + 
   scale_x_date(expand = c(0,0)) +
  labs(title = "Line Plot",
       subtitle = "theme_wombat()",
       caption = "Data: economics_long") +
  theme_wombat()

s2 <- ggplot(mtcars, aes(factor(carb),fill=factor(carb))) + 
  geom_bar() + 
  scale_y_continuous(expand = c(0,0)) + 
  scale_x_discrete(expand = c(0,0)) +
  labs(title="Bar Plot",
       subtitle = "theme_seagull()",
       caption = "Data: mtcars") +
  theme_seagull()

s3 <- ggplot(economics_long, aes(date, value01, colour = variable)) +
  geom_line() +
  scale_y_continuous(expand = c(0,0)) + 
  scale_x_date(expand = c(0,0)) +
  labs(title = "Line Plot",
       subtitle = "theme_seagull()",
       caption = "Data: economics_long") +
  theme_seagull()

bw1 <- ggplot(mtcars, aes(mpg,disp,color=factor(carb))) + 
  geom_point(size=3) + 
  labs(title="Scatter Plot",
       subtitle = "theme_bw()",
       caption = "Data: mtcars") +
  theme_bw()

bw2 <- ggplot(mtcars, aes(factor(carb),fill=factor(carb))) + 
  geom_bar() + 
  labs(title="Bar Plot",
       subtitle = "theme_bw()",
       caption = "Data: mtcars") +
  theme_bw()

bw3 <- ggplot(economics_long, aes(date, value01, colour = variable)) +
  geom_line() +
  labs(title = "Line Plot",
       subtitle = "theme_bw()",
       caption = "Data: economics_long") +
  theme_bw()

(s1 + w1 + bw1) / (s2 + w2 + bw2) / (s3 + w3 + bw3)
```

Side by side comparison of scatter plots, bar charts, and line charts between
the three themes can be found in figure \@ref(fig:theme-comparison). One of 
the differences between `dataviz` potential themes and `theme_bw()` is the size
of the text. The axis title, plot title, subtitle, and caption with 
`theme_seagull()` and `theme_wombat()` are easier to read in comparison to
`theme_bw()`. Secondly, the legend position is located on the bottom of the
plot as opposed to the right of the plot. The reason to the change in legend
position is to maximize the data-ink ratio of the plots. Putting the legend
on the right of the plot takes up the space that can be used to display the
underlying data. Another way of maximizing data-ink ratio is by reducing
the margin around the plot. As shown in the bar plots and line plots in figure \@ref(fig:theme-comparison), there are no extra space between the bars/lines 
with the axis lines in the plots with `theme_seagull()` and `theme_wombat()`.

One of the objectives of `theme_seagull()` is to have a theme that increases 
data-ink ratio as much as possible. In comparison to other two themes, 
`theme_seagull()` has the least amount of grid lines and borders around the
plot, making it to have the highest data-ink ratio. While there is no vertical 
grid lines, major horizontal grid lines are preserved as they aid readers to 
approximate and compare values between data points.

`theme_wombat()` is developed to be the middle ground between `theme_seagull()`
and `theme_bw()`. In comparison to the `ggplot2` built-in theme, this theme
is less cluttered as it does not have minor grid lines on both axis. Unlike
`theme_seagull()`, however, this theme retains major vertical grid lines.

### The Chosen Theme

As stated above, `theme_seagull()` has the least clutters between the two 
potential themes. However, `theme_wombat()` is likely to assist readers
to perform comparative exercise more accurately comparative to `theme_seagull()`
while still maintaining a high level of data-ink ratio. Using the scatter plot 
example in figure \@ref(fig:theme-comparison), the major vertical grid lines 
from `theme_wombat()` aids readers estimate the value of data points along the
x axis. This shows that `theme_wombat()` is a good compromise of minimizing 
clutters as well as allowing readers to extract information from the graph as 
much as possible.

\newpage

# Choosing Color Palletes

In addition to themes, color also plays an important role in data 
visualizations. Proper use of color in a visualization enhances the quality
of information being presented [@zeileis2009escaping]. One of the most important
use of color in data visualizations is to differentiate one element from one
another [@stone2006choosing]. However, a poor choice of color palette can impede 
viewers' ability to effectively extract information from the visualization. 

With a simple visual features, some objects can be easily recognised compared
to other objects without requiring attention. The target object will "pop out" 
in field of vision regardless how many distractors surrounding it 
[@treisman1985preattentive]. The use of color in data visualisation can facilitate
this "pop-out" as it makes it can make some elements to be easier to be found.

```{r pop-out, fig.cap="It is easier to find the odd ones by using color"}
set.seed(20211028)

x <- rnorm(20)
y <- rnorm(20)
color <- c("red","red","blue","red","blue","red","red","red","red","red","red",
           "blue","red","red","red","red","red","red","red","red")
pop_out <- data.frame(x,y,color)

p1 <- pop_out %>% ggplot(aes(x = x,
                             y = y,
                             color = color)) +
  geom_point() +
  theme_wombat() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  ggtitle("Color")

p2 <- pop_out %>% ggplot(aes(x = x,
                             y = y,
                             shape = color)) +
  geom_point() +
  theme_wombat() +
  theme(legend.position = "none",
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank()) +
  ggtitle("Shape")

p1 + p2
```

Figure \@ref(fig:pop-out) contains three "odd" observations that are different 
than the rest. To represent this, the first panel uses color to differentiate 
the different kinds of observations. The three observations are colored red, and
the rest are blue. On the other hand, the second panel uses shape to represent
this feature. Circle is used to represent the "odd" observations, and triangle is
used for the rest. While readers have no problem in identifying the three 
observations in the second panel, it is much easier to find these observations 
on the first panel. This shows that finding the observations of interest in the 
first panel requires significantly lower level of attentions compared to the 
second panel.

In addition to assisting viewers in differentiating one attribute to another, 
colors can also be used to evoke viewers' emotions. Emotion is important
in visualizations for story telling, communicative intent, and engagement
[@bartram2017affective].  It was also found that emotions can significantly 
improve many cognitive processes such as attention, learning, decision making, 
and problem-solving [@harrison2012exploring].

## Color Spaces

To effectively use colors in visualizations, it is imperative to understand
how human color perception works. It was theorized that human color vision evolved 
in three distinct stages [@ihaka2003colour]: 

1. Perception of light and dark (monochrome)
2. Perception of yellow/blue contrast
3. Perception of green/red contrast

As a result of these three color axes, colors are naturally described as 
locations in a three dimensional space. While these color axis provide 
physiological description, human perception of colors do not corresponds to
these axes. Human perception of color corresponds to the polar coordinates
of the color coordinate (yellow/blue and green/red), and a dark/light axis 
[@zeileis2006choosing]:

1. Hue: dominant wavelength
2. Chroma: colorfulness (color intensity)
3. Luminance: brightness (brightness)

Unfortunately, not all software packages use color space was developed based on
human perception. For example, the most common color map used by software 
packages, the Red-Green-Blue (RGB) [@stauffer2015somewhere], was primarily 
developed to correspond with color generation on computer screens rather than 
to correspond with human perception of color [@zeileis2009escaping]. 
Zeileis, Hornik, and Murrell further stated that it is practically impossible
for a human to control the perceptual properties of a color within this color 
model as there is no single dimension that relates to hue or brightness of a 
color

One popular alternative to RGB system is Hue-Saturation-Value (HSV). This model
is a simple transformation of RGB system and aims to capture the dimensions of 
human visual perceptions [@smith1978color]. Unfortunately, the result of mapping 
HSV's three dimensions to three dimensions of human perceptions is very poor
[@zeileis2006choosing]. Moreover, it is fairly difficult to find a set of colors 
that are in harmony in HSV color space [@zeileis2006choosing]. Furthermore, HSV encourages 
the use of colors that are highly saturated. While these colors are good to 
attract viewers' attention, it is difficult to look at for a long time 
[@zeileis2009escaping].

HCL, which is short for Hue-Chroma-Luminance, was developed to overcome many 
problems encountered by other color spaces. This color space is developed
based on the three color attributes that corresponds to human perceptions:
hue, chroma, and luminance [@sarifuddin2005new]. It was found that the three
dimensions from this color space matches human visual perception well 
[@zeileis2019colorspace]. In general, HCL color space can be used to generate 
three types of color scales: qualitative, sequential, and diverging.

### Qualitative Palettes

This color palette is designed for coding categorical information where the 
categories do not have particular ordering [@zeileis2019colorspace]. In other
words, this color palette will give the same perceptual weight to each category
so that there will be no perception of one group to be larger or more important
than any other one [@zeileis2006choosing]. This color palette is commonly 
applied to bar plots or box plots.

### Sequential Palettes

Sequential pallets are used for coding ordered/numerical information that span
in a certain interval [@zeileis2006choosing]. The low values of the interval
are considered to be uninteresting, whereas high values are considered to be 
interesting [@zeileis2019colorspace]. Low values are usually depicted by lighter
colors, while darker colors are used for high values [@harrower2003colorbrewer].
This palette is usually used on heatmaps.

### Diverging Palletes

Similar to sequential palette, diverging pallets are designed for coding ordered/
numerical information that ranges in an interval. The difference between 
diverging and sequential palletes is that diverging pallets include a neutral
value [@zeileis2006choosing]. This palette should be used when an important data
class or break points need to be highlighted [@harrower2003colorbrewer]. Harrower 
and Brewer further stated that this critical data class or break points is 
highlighted by a change in in hue and luminance, and should represent a critical 
value of the data (e.g. mean, median, or zero). For instance, the average GDP
per-capita in a choropleth map can be emphasized (i.e. set as neutral value) so 
that the values above and below the average can be shown in different hues.

## Color palette of choice

In addition to plot themes, this project also contributes two color schemes for 
each of the three color palettes type mentioned in the previous section. 
`colorspace` and `pals` packages are going to be used to build the color palettes 
and assess them. The chosen color palettes need to be assessed to ensure
that they achieve their intended purpose. For example, qualitative color palette
needs to avoid implicit ordering as this is intended for showing coding categorical 
information. In addition, it needs to be ensured that people with color vision
impairment are able to distinguish the different colors in the chosen palettes.
`pal.safe()` function from `pals` package allows us to see how color palettes
are seen by deutan, protan, and tritan color-blind.

The methodology of choosing color palettes that are not only accessible but also 
enhance the quality of visualizations will be discussed in this section. Firstly, the
report will explore common strategies used in choosing qualitative, sequential, 
and diverging palettes. Then, examples of palettes made from these strategies will
be assessed. Lastly, the report will show how color schemes for 
qualitative, sequential and diverging palettes are chosen for the project.

### Chosen Qualitative Palettes

```{r color-blind, fig.height=8, fig.cap="Color palettes as seen by people with vision impairment"}
quokka_palette <- c("#8C01E6", "#93B1EB", "#AB313D", "#D6CA24")
quoll_palette <- c('#8D5698', '#EFD36F', '#247A3E', '#17B0A8')
seal_palette <- colorspace::sequential_hcl(7,
                                              h = c(0, -140),
                                              c = c(80,40),
                                              l = c(40, 75),
                                              power = c(1,1))
snake_palette <- colorspace::sequential_hcl(6,
                                              h = c(250, 90),
                                              c = c(40, 55),
                                              l = c(33,98),
                                              power = c(0.5,1),
                                              rev = TRUE)
dingo_palette <- colorspace::diverge_hcl(7,
                                           h = c(260, 0),
                                           c = 80,
                                           l = c(30, 90),
                                           power = 1.5)
dolphin_palette <- colorspace::diverging_hcl(n = 7,
                                             h = c(265, 45),
                                             c = 80,
                                             l = c(70, 95),
                                             power = c(0.5, 2),
                                             rev = TRUE)

# colorblind filter
par(mfrow=c(4,2))
pal.safe(qualitative_hcl(4, palette = "harmonic"), main = "Harmonic Palette")
pal.safe(quokka_palette, main = "Quokka Palette")
pal.safe(quoll_palette, main = "Quoll Palette")
pal.safe(seal_palette, main = "Seal Palette")
pal.safe(snake_palette, main = "Snake Palette")
pal.safe(dingo_palette, main = "Dingo Palette")
pal.safe(dolphin_palette, main = "Dolphin Palette")
```

```{r harmonic, fig.cap="HCL color spectrum plot for Harmonic palette"}
specplot(qualitative_hcl(4, palette = "harmonic"), main = "Harmonic Palette")
```

One common strategy to choose this type of palette is by keeping chroma and luminance
relatively constant, and hue varied; making all colors in the palette to balance 
towards the same grey [@zeileis2009escaping]. While this strategy is relatively
simple, keeping chroma and luminance fixed may not be the best approach to 
choose color palette that are accessible.

Figure \@ref(fig:harmonic) shows an example of a qualitative color palette from 
`colorspace` package called harmonic. As shown in the figure, chroma
and luminance remain constant, while the hue varies across the four colors.
It is fair to say that this color set works relatively well to show categorical 
information as it does not give implied ordering across different categories.
This method, however, does not have the same result for people with visual
impairment. It can be seen in figure \@ref(fig:color-blind) that people 
with color-blindness will have difficulty in distinguishing some colors from 
harmonic palette. For instance, there are three colors that look similar as seen 
by people with tritan color-blindness. Moreover, this color palette results in 
implied ordering of different data group for people with visual impairment.

```{r quokka, fig.cap="HCL color spectrum plot for Quokka palette"}
specplot(quokka_palette, main = "Quokka Palette")
```

```{r quoll, fig.cap="HCL color spectrum plot for Quoll palette"}
specplot(quoll_palette, main = "Quoll Palette")
```

The two qualitative color palettes developed for the project are called 
`quokka_palette()` and `quoll_palette()`. Figure \@ref(fig:quokka) and figure 
\@ref(fig:quoll) shows the HCL color spectrum plot for these palettes. In
contrast with the Harmonic Palette, Quokka and Quoll palettes have differing 
hues, chroma and luminance across all colors. As a result, each color in
these palettes can easily be distinguished by people with deutan, protan, and 
tritan color-blindness (see figure \@ref(fig:color-blind)). However, removing
implied ordering for qualitative palettes while ensuring that the palettes
are accessible proves to be difficult. Figure \@ref(fig:color-blind) shows that
while Quokka and Quoll palettes are better compared to Harmonic palette in
not showing implied ordering for people with visual impairment, there are still 
some colors that look "paired".

### Chosen Sequential Palettes

The strategy to choose a sequential palette is different to that of qualitative
palette. One easy way to create a sequential palette is by choosing two values 
for hue (h1 and h2), chroma (c1 and c2), luminance (l1 and l2), and power 
transformations (p1 and p2) for chroma and luminance [@zeileis2006choosing].
These values of hues, chroma, luminance, and power transformation can be used
as inputs to `sequential_hcl` function to produce a set of sequential palette.

```{r seal, fig.cap="HCL color spectrum plot for Seal palette"}
specplot(seal_palette, main = "Seal Palette")
```

```{r snake, fig.cap="HCL color spectrum plot for Snake palette"}
specplot(snake_palette, main = "Snake Palette")
```

The two sequential palettes created for this project are `seal_palette()` and
`snake_palette()`. As shown in figure \@ref(fig:seal) and figure \@ref(fig:snake),
these palettes use multi-hue schemes as opposed to single-hue schemes. Multi-hue
palettes were chosen for the project as it provides better contrast between different
classes compared to single-hue sequential palette [@harrower2003colorbrewer].
Moreover, multi-hue sequential palettes avoids the use of extreme colors, as well
as allow viewers to better discriminate the middle colors [@zeileis2019colorspace]. 
In addition, figure \@ref(fig:color-blind) shows that people with deutan, protan, 
and tritan color blindness are able to distinguish the colors in the palette.
This means that color palettes will allow readers with color vision deficiency
to see implied order and how the data ranges from low to high values.

### Chosen Diverging Palettes

The approach in choosing diverging palettes is similar to choosing sequential
palettes. Diverging palettes can be built by choosing two values for
hues, chroma, luminance, and power transformation for chroma and luminance 
[@zeileis2006choosing]. Function `diverge_hcl()` is used to create a diverging 
palette based on these values, and it will automatically choose the neutral color.

```{r dingo, fig.cap="HCL color spectrum plot for Dingo palette"}
specplot(dingo_palette, main = "Dingo Palette")
```

```{r dolphin, fig.cap="HCL color spectrum plot for Dolphin palette"}
specplot(dolphin_palette, main = "Dolphin Palette")
```

One condition needs to be met when choosing a color pair for a diverging
palette: the color pair should not create color confusions: naming confusions, 
color vision impairment, and simultaneous contrast confusion 
[@brewer1996guidelines]. Naming confusion happens when people use two 
basic color names for the same color. On the other hand, simultaneous contrast
refers to a phenomena where colors are affected by their opposite. For instance,
a color will appear lighter when surrounded by a dark color, and a color will
appear warmer when surrounded by a cool color [@brewer1992review]. This can be
prevented by avoiding selecting opposite color pairs for diverging palettes
[@brewer1996guidelines].

Dingo and Dolphin palettes are diverging palettes developed for the project.
These palettes are shown in figure \@ref(fig:dingo) and \@ref(fig:dolphin).
Red/blue pairs and orange/blue hue pairs are chosen for Dingo and Dolphin
palettes respectively as these pairs create no naming confusions and 
simultaneous contrast confusions [@brewer1996guidelines]. Moreover, these 
palettes create no color confusion due to color vision impairment as colors in 
Dingo and Dolphin palettes can be easily distinguished by people with color-blindness
(shown in figure \@ref(fig:color-blind)).

# R Package

One of the objectives of this project is to create a solution for authors and
editors of the research book to create visualizations that are readable, 
cohesive, and accessible. The solution needs to be easy to use so that it
does not disrupt authors' workflow. Tr po achieve this, R codes necessary
to create the chosen `ggplot2` theme and color schemes discussed in the previous
two sections will be bundled in an R package called `dataviz`. R packages are 
extensions to R that can comprise of codes, data, documentations, tests 
[@wickham2015r]. 

One of the reasons of creating an R package for this project is because of ease
of use. Codes, documentation, and tests in an R package are built in a 
standardized format so that any R users can install it [@wickham2015r]. In 
addition, it is fairly simple to create an R package. Packages such as `Roxygen2`,
`devtools`, and `testthat` provide great assistance in documenting and testing
the package. Another benefit of using an R package is that updates to the themes
and color schemes are not time consuming. Any updates to themes and color schemes
simply need to be applied to the codes in R package; there is no need to update
each plots in the book. This shows that using R package will allow us to achieve 
one of the aim of the project: to provide solutions that cause minimal 
disruption to authors' work.

R packages are typically stored in in a repository. Two of the most common
repositories for R packages are Comprehensive R Archive Network (CRAN) and
GitHub. For this project, the `dataviz` package will be stored in GitHub.
GitHub is a popular choice for a repository as it has the integration with git, 
a version control software. Version control records changes applied to the 
package to a file. This allows user to recall specific versions of the package
at later time [@noauthor_git_nodate].

## R Package Components

There are several components inside an the `dataviz` package: R codes, package 
metadata, documentation, vignettes, tests, and Namespace. 

R project 

R codes within an R package is stored in `R/` directory.

P4: R package components:
* R packages contain R codes, package metadata, documentation, vignettes, 
tests, and Namespace.

## How `dataviz` was created

## Testing

# Conclusion (300)
